---
title: "Datasets: Description and Curation Protocol"
author: "Luis Damiano, Jarad Niemi"
date: "`r Sys.Date()`"
output: 
  rmarkdown::pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
colorlinks: true
urlcolor: blue
linkcolor: blue
citecolor: red
header-includes:
  - \usepackage{hyperref}
vignette: >
  %\VignetteIndexEntry{STRIPS yield datasets: description and curation Protocol}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

\hypersetup{
  colorlinks   = true,  % Colours links instead of ugly boxes
  urlcolor     = blue,  % Colour for external hyperlinks
  linkcolor    = blue,  % Colour of internal links
  citecolor    = red    % Colour of citations
}

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo     = FALSE,
  cache    = FALSE,
  comment  = "#>"
)

suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(shapefiles))
suppressPackageStartupMessages(library(STRIPSyield))
suppressPackageStartupMessages(library(reshape2))

describe_shapefile <- function(shape) {
  describe_column <- function(shape, columnName) {
    column <- shape$dbf$dbf[, columnName]
    data.frame(
      name     = columnName,
      class    = class(column)[1],
      nFactors = if (is.factor(column)) { length(unique(column)) } else { NA },
      min      = if (is.factor(column)) { NA } else { min(column)  },
      mean     = if (is.factor(column)) { NA } else { mean(column) },
      max      = if (is.factor(column)) { NA } else { max(column)  },
      example  = trimws(as.character(column[1]))
    )
  }

  do.call(
    rbind,
    lapply(colnames(shape$dbf$dbf), describe_column, shape = shape)
  )
}

describe_yield <- function(yield) {
  describe_column <- function(yield, columnName) {
    column <- yield[, columnName]
    data.frame(
      name     = columnName,
      class    = class(column)[1],
      nFactors = if (is.factor(column)) { length(unique(column)) } else { NA },
      min      = if (is.factor(column)) { NA } else { min(column)  },
      mean     = if (is.factor(column)) { NA } else { mean(column) },
      max      = if (is.factor(column)) { NA } else { max(column)  },
      example  = trimws(as.character(column[1]))
    )
  }

  do.call(
    rbind,
    lapply(colnames(yield), describe_column, yield = yield)
  )
}
```

# Accessing data

## How to access the datasets

The followings are the datasets contained in this R Package:

* **Point-level yield monitor data**:
    * **R data frames**: after the package is loaded, two objects `yield` and
`yieldExtra` containing point-coordinate level yield data frames become
available. These are described in Table \ref{tab:yieldExtra-description} and
displayed in Figures \ref{fig:data-visualization-soybeans} and
\ref{fig:data-visualization-corn}.
    * **Shapefiles**: datasets in shapefile format are available as discussed in
Section \ref{sec:curation-protocol}.
* **PNAS paper**: to access the dataset from the PNAS paper [1], load the 
package and run `pnas_data()`.

```{r yieldExtra-description, eval = FALSE}
dYield <- describe_yield(yieldExtra)
units  <- c(
  NA, NA, NA, "Hectare", NA, "Hectare", NA, "% (hundreds)", NA, 
  "% (hundreds)", NA, NA, "Feet", "Integer", NA, 
  NA, NA, "LatLong", "LatLong", "UTM", "UTM", NA, "Feet", "MPH", "Degrees", 
  "Feet", "Seconds", "Pounds per second", "% (hundreds)", 
  "Bushels per acre (dry)"
)
extra <- ifelse(colnames(yieldExtra) %in% colnames(yield), "Yes", "No")
colNames <- c(
  "Name", "Class", "# of levels", "Min", "Mean", "Max", "Example"
)

tab <- cbind(dYield, units = units, "extra only" = extra)
colnames(tab) <- c(colNames, "Measurement unit", "yieldExtra only")

k <- knitr::kable(
  tab,
  format = "latex", booktabs = TRUE,
  label = "yieldExtra-description",
  caption = "Structure and content of the yield and yieldExtra datasets.",
  digits = 2, format.args = list(scientific = FALSE)
)

kableExtra::kable_styling(k, latex_options = c("scale_down"))
```

## Terminology

The dataset adheres to the terminology used in [2] to describe the experimental
design.

* __Site__: three locations within the Neal Smith National Wildlife Refuge
(NSNWR) in central Iowa (namely Basswood, Interim, and Orbweaver).
* __Blocks__: there are four blocks (namely BasswoodA, BasswoodB, Interim, and 
Orbweaver).
* __Watershed__: twelve experimental units (Basswood 1 to 6, Interim 1 to 3,
Orbweaver 1 to 3).
* __Field__: the portion of the whatershed planted with either row crops or 
perennial vegetation.
* __Treatment__: four watershed-scale treatments having different proportions 
and topographic positions of PFS (no PFS, 10% PFS at toeslope position, 10% PFS 
distributed on toe and contour strips, and 20% PFS distributed on toe 
and contour strips).
* __Coordinate point__: each of the spatial coordinate units with recorded 
information (number and position of the points vary per year and treatment).

# Curation protocol

\label{sec:curation-protocol}

## Folder structure

In STRIPSyield v0.2.0, the datasets are stored in the folders: 
`data-raw\source\YYYY-site.ext`.

  - `legacy`: CSV existing in STRIPSYield v0.1.1 that were produced by an
  unspecified methodology.
  - `original`: shapefiles as received from the Coop. Besides modifying the file
  name, file structure and content are kept intact.
  - `curated`: new shapefiles originating from the curation protocol. Note that
  this process modifies both the datasets structure and content.

## Curation protocol

Because not all the datasets have the same structure and measurement units, we
create a curation protocol. We identify two patterns in the data sources, namely
Template I (2007-2010 and 2012) and Template II (2013-2019 and 2011). We read
the shapefiles from the `original` folder, apply the modifications mentioned
below, and store the new shapefiles in the `curated` folder. These editing rules
may be broadly classified into five actions:

  - __Rename__: we modify the name of the variable but not the content.
  - __Reformat__: we modify the name of the variable and the content
  (e.g. change of measurement unit).
  - __Drop__: we discard some content if it is not present in every shapefiles
  across the years and sites.

Although keeping both the original and the curated shapefiles result in
significant storage redundancy, this procedure guarantees that no original data
is lost in the process.

### Naming convension

File naming convention:

* data-raw/yield_original/YYYY-site.ext
* data-raw/yield_curated/YYYY-site.ext
* Note that we use hyphen to separate words, and site names are lowercase.

Column naming convention:

* Use camelCase (e.g. prairiePosition). Note that the starting letter is
lowercase.
* No measurement units in the column names. For measurement units,
see this vignette.

Data structure convention:

* All strings as factors.
* All strings start with uppercase. (ex. Soybeans, Orbweaver).
* Dates and timestamps are Date and POSIXct objects respectively.
* Use `NA` for missing data.

### Original shapefiles (2007-2010, 2012)

The PROJ4 string defining the CRS of the coordinates recorded in these
shapesfiles is `"+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84
+towgs84=0,0,0"`. Since the 2007-2010 and 2012 shapefiles are consistent,
`2009-basswood` original shapefile structure is described in Table
\ref{tab:2009-basswood-description}. To homogenenize measurement units, we
rescale the columns `distance` and `swath` width from inches to foot.

```{r 2009-basswood-description, eval = FALSE}
shape2009  <- read.shapefile("../data-raw/yield_original/2009-basswood")
dShape2009 <- describe_shapefile(shape2009)

tab <- dShape2009
colnames(tab) <- colNames
k <- knitr::kable(
  tab, 
  format = "latex", booktabs = TRUE,
  digits = 2, format.args = list(scientific = FALSE),
  caption = "Structure and content of the Basswood 2015 original shapefile.",
  label = "2009-basswood-description"
)

kableExtra::kable_styling(k, latex_options = c("scale_down"))
```

### Original shapefiles (2013-2019, 2011)

The PROJ4 string defining the CRS of the coordinates recorded in these
shapesfiles is `"+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84
+towgs84=0,0,0"`. Since the 2013-2019 and 2011 shapefiles are consistent, the
`2015-basswood` original shapefile structure is described in Table
\ref{tab:2015-basswood-description}.

```{r 2015-basswood-description, eval = FALSE}
shape2015  <- read.shapefile("../data-raw/yield_original/2015-basswood")
dShape2015 <- describe_shapefile(shape2015)

tab <- dShape2015
colnames(tab) <- colNames
k <- knitr::kable(
  tab, 
  format = "latex", booktabs = TRUE,
  digits = 2, format.args = list(scientific = FALSE),
  label = "2015-basswood-description",
  caption = "Structure and content of the Basswood 2015 original shapefile."
)

kableExtra::kable_styling(k, latex_options = c("scale_down"))
```

### Curated shapefile (all years)

The PROJ4 string defining the CRS of the coordinates recorded in these
shapesfiles is `"+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84
+towgs84=0,0,0"` (no projections were needed). As the shapefile structure and
content does not vary across the years and sites, the structure described in Table
\ref{tab:curated-shapefile-description} is valid for every curated shapefile.

```{r curated-shapefile-description, eval = FALSE}
shape2015  <- read.shapefile("../data-raw/yield_curated/2015-basswood")
dShape2015 <- describe_shapefile(shape2015)
units      <- c(
  NA, NA, "Feet", "Integer", "Integer", NA, "WGS84", "WGS84", "Feet above sea", 
  "Miles per hour", "Arc degrees", "Feet", "Seconds", NA, "Pounds per second", 
  "% (hundreds)", "Dry bushels per acre"
)

tab <- cbind(dShape2015[, 1:2], units)
colnames(tab) <- c("Name", "Class", "Measurement unit")
k <- knitr::kable(
  tab,
  format = "latex", booktabs = TRUE,
  digits = 2, format.args = list(scientific = FALSE),
    caption = "Structure a curated shapefile.",
  label = "curated-shapefile-description"
)

kableExtra::kable_styling(k)
```

To build our consolidated shapefiles, we decided to keep only those variables
recorded for every site and year (i.e. columns that were present in every source
file). The only exceptions is `direction`, available for years 2013-2015 and
2011 only, which we kept as partial information may be relevant for our future
research.

# References

[1] Lisa A. Schulte, Jarad B. Niemi, Matthew J. Helmers, Matt Liebman, J. G.
Arbuckle, David E. James, Randall K. Kolka, Matthew E. O’Neal, Mark D. Tomer,
John C. Tyndall, Heidi Asbjornsen, Pauline Drobney, Jeri Neal, Gary Van Ryswyk,
and Chris Witte (2017). “Prairie strips improve biodiversity and the delivery of
multiple ecosystem services from corn-soybean croplands” Proceedings of the
National Academy of Sciences, 114(42), 11247-11252.
([url](http://www.pnas.org/content/114/42/11247.short))

[2] Xiaobo Zhou, Matthew J. Helmers, Heidi J. Asbjornsen, Randy Kolka, and Mark
D. Tomer (2010). "Perennial filter strips reduce nitrate levels in soil and
shallow groundwater after grassland-to-cropland conversion" Journal of
environmental quality, 39(6), 2006-2015.

\clearpage

# Data visualization

\clearpage

```{r data-visualization}
plot_grid <- function(df, title = NULL) {
  p <- ggplot(df) +
  geom_violin(aes(x = year, y = value), draw_quantiles = c(0.5)) +
  facet_grid(
    rows = vars(variable),
    cols = vars(site),
    scales = "free"
  ) +
  labs(
    title = title,
    x     = "Year",
    y     = "Values"
  ) +
  theme_bw() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.x        = element_text(angle = 90, hjust = 1)
  )
  
  suppressWarnings(print(p))
}

keys <- c("site", "year", "crop", "record")
cols <- c(
  "swath", "elevation", "speed", "distance",
  "cycle", "flow", "moisture", "yield"
)

yieldExtraLong <- melt(
  data = yieldExtra,
  id.vars = keys,
  measure.vars = cols
)
```

```{r, out.width = "\\textwidth", fig.width = 9, fig.height = 11, fig.cap = "\\label{fig:data-visualization-soybeans} Visualization of some selected `yieldExtra` data frame variables.", warning = FALSE}
plot_grid(
  df = yieldExtraLong[
    yieldExtraLong$site != "OffBounds" & 
    yieldExtraLong$crop == "Soybeans", 
  ],
  title = "Soybeans"
)
```

```{r, out.width = "\\textwidth", fig.width = 9, fig.height = 11, fig.cap = "\\label{fig:data-visualization-corn} Visualization of some selected `yieldExtra` data frame variables.", warning = FALSE}
plot_grid(
  df = yieldExtraLong[
    yieldExtraLong$site != "OffBounds" & 
    yieldExtraLong$crop == "Corn", 
  ],
  title = "Maize"
)
```

\clearpage

# Tables

\clearpage

```{r yieldExtra-description, eval = TRUE}
```

```{r 2009-basswood-description, eval = TRUE}
```

```{r 2015-basswood-description, eval = TRUE}
```

```{r curated-shapefile-description, eval = TRUE}
```
